---
title: "Analysis of The Safety of Self Driving Cars"
author: "Shreya Tarat"
date: "2024-11-11"
output: 
  html_notebook:
    fig_height: 7
    fig_width: 10
---
# Research Question:
Are cars equipped with Automated Driving Systems safer than the normal vehicular cars on road?

## Become acquainted with your data sources: 
###	Where did you find them? 
I found my two primary datasets from Kaggle.com

###	Who collected/maintains them? 
*	The Data is from the National Highway Traffic Safety Administration for Automated car crashes 
* The second dataset is from a Countrywide Traffic accident reports dataset published by Arxivs. 

###	When and why were they originally collected? 
*	Automated car crashes dataset was collected in 2021. The HHTSA issued a directive requiring manufacturers and fleet operators to report certain crashes involving either advanced driver assistance or higher-level “automated driving systems.
*	Vehicular accidents dataset was collected from Feb 2016-2023 for numerous applications, such as real-time car accident prediction, studying car accident hotspot locations, casualty analysis etc.

###	What does a case represent in each data source, and how many total cases are available?
* Each case describe each report and include information about the car, location, circumstances, and injury level in both the datasets
* Automated car crashes have 436 cases while Vehicular dataset consists of 7728394 cases.

###	What are some of the variables that you plan to use? 
Severity of accident, model of the car, 


### Load packages
```{r}
library(readr)
library(dplyr) 
#Using data.table library to use 'fread'. I found this resource while researching online and is used to load large datasets. This is much faster than read_csv 
library(data.table)
library(tidyverse)
```
### Read data
```{r}
ads_data_small_sample <- ads_data[sample(nrow(ads_data), 100), ]

# Save the smaller dataset to a new CSV file
write.csv(ads_data_small_sample, "ADS_Incident_Reports_Small.csv", row.names = FALSE)
car_accidents <- fread("US_Accidents_March23.csv")
head(car_accidents)
```

### Explore intuition related to the research question

The graph below shows the severity statistics of the ADS driving accident report. It shows that most accidents have no injuries reported
```{r}
ggplot(ads_data) +
  aes(x = `Highest Injury Severity`) +
  geom_bar(fill = "#112446") +
  theme_minimal()
```
The graph below shows the severity of the car accident reports. It also shows that most accidents here cause little traffic obstruction
```{r}
ggplot(car_accidents) +
  aes(x = `Severity`) +
  geom_bar(fill = "#112446") +
  theme_minimal()
```
### Describe preliminary observations and intuition about the research question
*	We observe ADS cars have less severe injuries, giving them a safety advantage as they react faster than humans
*	I plan on adding demographics and environmental factors to observe how they performed during these conditions. 
*	Frequency of ADS car accidents are less, which suggest better safety

### TWO Data Sources (at least): 

* Primary data 
  + https://www.kaggle.com/datasets/michaelbryantds/automated-driving-crashes
* Other data sources 
  + https://www.kaggle.com/datasets/sobhanmoosavi/us-accidents
  + https://arxiv.org/abs/1906.05409	
