---
title: "Analysis of The Safety of Self Driving Cars"
author: "Anonymous"
date: "2024-11-11"
output: 
  html_notebook:
    fig_height: 7
    fig_width: 10
---
# Research Question:
How do geographic and environmental characteristics effect the severity and the frequency of accidents in the United States?

# Relevance
Understanding the relationship between geographic and environmental factors and traffic accidents is critical as will help in priortizing safety interventions and guide infrastructure updates, like better road contions, lighting etc. 


## Primary Source: 
The dataset is from a Countrywide Traffic accident reports dataset published by Arxivs.

## Secondary Source:
The dataset is ZipDemography which is taken from dcData package. 

## Challenges faced
The dataset was very large and I needed a way to decrease the query times without effecting the quality of the analysis. I decided to use the fread function to read the csv instead of read_csv since it is much faster to read. Also I created a smaller dataset of 100,000 random rows instead of the original 7 million. This improved the query times and still gave me a good sample to work with.

## Load packages and data
```{r}
library(readr)
library(dplyr) 
#Using data.table library to use 'fread'. I found this resource while researching online and is used to load large datasets. This is much faster than read_csv 
library(data.table)
library(tidyverse)
library(dcData)
data("ZipDemography")
cleaned_accident_data <- fread("US_Accidents_March23_100K.csv")
head(cleaned_accident_data, 4)
```
## Cleaning the data
Here we remove the rows with missing information about Severity, Start_Time, Zipcode, Weather_Condition, Temperature(F), Visibility(mi), and Severity. We also need to convert the Start Time into a POSIX date object.
```{r}
cleaned_accident_data <- cleaned_accident_data %>%
  filter(
    !is.na(Severity),
    !is.na(Start_Time),
    !is.na(Zipcode),
    !is.na(Weather_Condition),
    !is.na(State)
  ) %>%
  mutate(Start_Time = ymd_hms(Start_Time)) %>%
  select(Severity, Start_Time, Zipcode, Weather_Condition, State)
```
# Does environment matter?
Here I want to check if weather conditions can tell us anything about car accidents.
```{r}
weather_summary <- cleaned_accident_data %>%
  group_by(Weather_Condition) %>%
  summarise(
    count = n(),
    avg_severity = mean(Severity), # Averaging the severity score for weather condition
  ) %>%
  arrange(desc(count)) %>%
  slice_head(n = 15)  # Taking only top 15 into account because conditions with less data do not tell us much
```
## Graph
The graph below shows that most accidents occur on fair conditions. I believe this is the cause of people driving at higher speeds during good conditions. Leading people to make more mistakes and slow reaction times make it hard to avoid accidents.
```{r}
ggplot(weather_summary, 
    aes(x = reorder(Weather_Condition, count), y = count)) +
    geom_bar(stat = "identity", aes(fill = avg_severity)) +
    coord_flip() +
    scale_fill_viridis_c() +
    theme_minimal() +
    labs(
      title = "Accident Frequency and Severity by Weather Condition",
      x = "Weather Condition",
      y = "Number of Accidents",
      fill = "Average\nSeverity"
    ) 
```

# What hour of the day do accidents occur the most?
Here we will check if the hour of the day could tell us about accident frequency. I expect to see more accidents during peak hours.
```{r}
cleaned_accident_data <- cleaned_accident_data %>%
  mutate(hour = hour(Start_Time))
time_weather_summary <- cleaned_accident_data %>%
    group_by(hour) %>%
    summarise(
      count = n(),
      avg_severity = mean(Severity)
    )
```
## Graphing it
We see here that the hypothesis was correct. The peak hours which are during people leaving for work (6-9) and people coming back from work (3-6) has the most number of accidents. The severity of the accidents do not change a lot.
```{r}
ggplot(time_weather_summary, 
    aes(x = hour, y = count)) +
    geom_bar(stat = "identity", aes(fill = avg_severity)) +
    scale_fill_viridis_c() +
    theme_minimal() +
    labs(
      title = "Accident Frequency and Severity by Hour",
      x = "Hour",
      y = "Number of Accidents",
      fill = "Average\nSeverity"
    )
```

# Does geography matter?
Here I want to analyze if the location, grouped by states can tell us about accident frequency or severity.
```{r}
# Getting data from ZipDemography
zip_data <- ZipDemography %>%
  filter(!is.na(Totalpopulation)) %>% # Remove rows not containing total population
  select(Totalpopulation, ZIP, Meantraveltimetoworkinminutespopulation16yearsandolder) # Only select required fields

# Removing '-' from zipcodes in accidents data. For example 123456-4123 -> 123456
cleaned_accident_data <- cleaned_accident_data %>%
  mutate(Zipcode = str_extract(Zipcode, "^[0-9]+")) # This matches a string with digits in the beginning until we see a non digit

# Join Accident Data with ZipDemography
accidents_with_zip_data <- cleaned_accident_data %>%
  left_join(zip_data, by = c(Zipcode = "ZIP"))

```
## Grouping by states
Here we group the data by states and count the total number of accidents.
```{r}
# Since zipcodes can be repeated in the joined accidents data, we need to first get the rows without duplicated zipcodes so we can use it to total population.
unique_zip_data <- accidents_with_zip_data %>%
  select(Zipcode, State, Totalpopulation) %>%
  distinct() # Removes duplicate ZIP codes

state_summary <- unique_zip_data %>%
  group_by(State) %>%
  summarise(
    total_population = sum(Totalpopulation, na.rm = TRUE) # Sum population for each zipcodes
  ) %>%
  left_join(
    accidents_with_zip_data %>% # Join it again so we can calculate the total number of accidents
      group_by(State) %>%
      summarise(
        total_accidents = n()
        ),
    by = "State"
  ) %>%
  arrange(desc(total_accidents))
top_10_states <- state_summary %>%
  slice_max(total_accidents, n = 10)
```
## Graphing Number of Accidents vs State
Here we see that California has the highest number of accidents and it is also the most populated state. Population density definitely has a big impact on number of accidents.
```{r}
ggplot(top_10_states, aes(x = reorder(State, total_accidents), y = total_accidents)) +
  geom_bar(stat = "identity", aes(fill = total_population)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  coord_flip() +
  labs(
    title = "Top 10 States with the Most Accidents",
    x = "State",
    y = "Number of Accidents",
    fill = "Total Population"
  )
```
# Travel Time Analysis
Here we will check if travel time to work effects the accidents.
```{r}
travel_time_data <- accidents_with_zip_data %>%
  group_by(Zipcode) %>%
  summarise(
    mean_travel_time = mean(Meantraveltimetoworkinminutespopulation16yearsandolder, na.rm = TRUE), # Calculating mean travel time
    total_accidents = n()
  )
```
## Graphing travel time
Here we see that a lot of accidents occur during the 25 minute work travel time. This indicates that as travel time increases the total number of accidents increases till a certain point (25 minutes in this case) and then total accidents seem to decrease. I believe that people with shorter travel time tend to make more mistakes because their commute might require them to navigate through more crowded places rather than highways.
```{r}
ggplot(travel_time_data, aes(x = mean_travel_time, y = total_accidents)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Accident Frequency vs. Mean Travel Time to Work",
    x = "Mean Travel Time (minutes)",
    y = "Total Accidents"
  ) +
  theme_minimal()
```
## Conclusion
We saw that geographic and environmental aspects does have impact on the number and severity of accidents.

Key Findings:
  + Most accidents occur during fair weather conditions. This could be attributed to drivers being less cautious, driving at higher speeds, and having slower reaction times.
  + Bad weather although causing less accidents, usually tend to have higher severity.
  + The peak hours of the day see a high number of accidents compared to other times.
  + States with high population like California and Texas see a high number of accidents.
  + Accidents are more frequent at 25 minutes of travel time, indicating short travel times play a role in accident frequency.


